---
title: "Bulk differential expression analysis of 'Neuroblastoma results' manuscript"
date: "`r Sys.Date()`"
author: "Anna Lovrics"
output:
  html_document:
    css: assets/rany_style.css
    df_print: paged
    highlight: tango
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
    number_sections: true
    code_folding: hide
---


```{r setup, include=F}
knitr::opts_chunk$set(message=FALSE, warnings=FALSE, fig.width=12, fig.height=5)
```

The code is under GNU GPL-3 license (https://www.gnu.org/licenses/gpl-3.0.html).

The objective of this R Markdown code is to reproduce bulk RNA-seq analysis
in the publication Tusnády et al.: 

```{r libraries}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
packages <- c(
  "ggplot2", "ggrepel",    # for some plots
  "tidyverse",            # Includes ggplot2, dplyr, tidyr, readr, purrr, forcats, tibble
  "limma",                # Differential expression analysis
  "edgeR",                # Normalization and DGEList
  "SummarizedExperiment", # Reading SummarizedExperiment assay data
  "ComplexHeatmap",       # Heatmap plotting & annotations
  "openxlsx",             # Writing Excel files
  "here",                 # Relative path management
  "AnnotationDbi",        # Database mapping
  "org.Hs.eg.db",         # Human gene annotation database
  "rtracklayer",          # Importing GTF annotation files
  "msigdbr", 
  "org.Hs.eg.db",
  "AnnotationDbi" 
)
# Load all packages quietly
invisible(lapply(packages, library, character.only = TRUE))
```

```{r folder structure}

Datafolder     <- here("Data")
# Subdirectories specifically isolated for this run's outputs
Figures   <- here(file.path("Figures", "reference_mapping"))
Data_calc <- here("Data_calculated")
Data_public <- here("Data_public")
Gencode_folder <- here("Gencode")
Results   <- here("Results")

# mainfolder <- "/home/anna/Anna/Enzimhome_local/Projects/Neuroblastoma"
datafolder <- here("Data") # here("Results_trimmed_321")
calcfolder <- here("Data_calculated")
if (!file.exists(calcfolder)) dir.create(calcfolder)
figfolder <- here("Figures")
if (!file.exists(figfolder)) dir.create(figfolder)
resfolder <- here("Results")
if (!file.exists(resfolder)) dir.create(resfolder)
```

Input for differential analysis is the count matrix, samples annotation and the design table. Gene annotation is also provided.
```{r read in input data}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

# Read in TSV input data uploaded to GEO
counts_file <- file.path(datafolder, "GSE335741_salmon.merged.gene_counts.tsv")
tpm_file    <- file.path(datafolder, "GSE335741_salmon.merged.gene_tpm.tsv")

# Read gene counts table
counts_df <- read.delim(counts_file, check.names = FALSE)

# Extract gene IDs / names if present in the TSV columns
# typical nf-core TSV format has 'gene_id' and 'gene_name' as first columns
if ("gene_name" %in% colnames(counts_df)) {
  gene_ids <- counts_df$gene_id
  counts <- as.matrix(counts_df[, -(1:2)])
  rownames(counts) <- gene_ids
} else {
  gene_ids <- counts_df[, 1]
  counts <- as.matrix(counts_df[, -1])
  rownames(counts) <- gene_ids
}

# Clean column names in count matrix to match downstream code
#colnames(counts) <- sub("_trimmed$", "", colnames(counts))

# Read sample metadata
annot_file <- file.path(datafolder, "samplesheet.csv")
metadata_scaled <- read.csv(annot_file)
samples_split <- strsplit2(metadata_scaled$sample, "_")
metadata_scaled$Cell_Line <- samples_split[,2]
metadata_scaled$Target <- paste(samples_split[,3], samples_split[,4], sep="_")
metadata_scaled$gRNA <- samples_split[,5]
sample_names <- metadata_scaled$sample

# metadata_scaled <- data.frame(Sample = sample_names)
# metadata_scaled$ID <- sub("_trimmed$", "", metadata_scaled$Sample)
# split_parts <- strsplit(metadata_scaled$ID, "_")
# 
metadata_scaled$Cell_Line <- factor(metadata_scaled$Cell_Line) 
metadata_scaled$Target <- factor(metadata_scaled$Target) 
metadata_scaled$gRNA <- factor(metadata_scaled$gRNA) 

# Re-level factors for a cleaner default contrast
metadata_scaled$Cell_Line <- relevel(metadata_scaled$Cell_Line, ref = "H1")
metadata_scaled$Target <- relevel(metadata_scaled$Target, ref = "STOP_MYCN")
metadata_scaled$gRNA <- relevel(metadata_scaled$gRNA, ref = "gRNA1")

# Ensure count matrix columns match metadata sample IDs
counts <- counts[, metadata_scaled$sample]

# Additive Design Matrix
design_additive <- model.matrix(~ Cell_Line + Target + gRNA, data = metadata_scaled)
rownames(design_additive) <- metadata_scaled$sample

# Read in genes annotation:
gene_annot_file <- file.path(Gencode_folder, "gencode.v49.primary_assembly.annotation.rds")
if (!file.exists(gene_annot_file)){
  gtf_file <- file.path(Gencode_folder, "gencode.v49.primary_assembly.annotation.gtf")
  gtf <- rtracklayer::import(gtf_file)
  
  # Extract gene-level annotations
  gene_annot <- gtf %>%
    as.data.frame() %>%
    filter(type == "gene") %>%
    select(gene_id, gene_name, gene_type,
           seqnames, start, end, strand)
  # save
  saveRDS(gene_annot, gene_annot_file)
} else {
  gene_annot <- readRDS(gene_annot_file)
}
```


Calculate and save normalised expression.
```{r calculate normalised expression}
# Convert count matrix to DGEList object
dge <- edgeR::DGEList(counts = counts)

# Perform filtering to remove genes with low counts
keep <- filterByExpr(dge, design_additive)
dge <- dge[keep, , keep.lib.sizes = FALSE]

# Apply TMM normalization
dge <- calcNormFactors(dge)

# Apply voom normalization
vv <- voom(dge, design_additive, plot = FALSE)

# Save & plot intermediate results
exprmx <- as_tibble(vv$E, rownames = "gene_id")
print(dim(exprmx))

# Merge with gene annotation
exprmx <- right_join(gene_annot, exprmx, by = "gene_id")
print(dim(exprmx))

exprfile <- file.path(resfolder, "logCPM.xlsx")
write.xlsx(exprmx, file = exprfile, colNames = TRUE, rowNames = FALSE)

# Load TPM counts table directly from Salmon GEO file
tpm_raw <- read.delim(tpm_file, check.names = FALSE)

if ("gene_name" %in% colnames(tpm_raw)) {
  tpm_ids <- tpm_raw$gene_id
  tpm_mat <- as.matrix(tpm_raw[, -(1:2)])
  rownames(tpm_mat) <- tpm_ids
} else {
  tpm_ids <- tpm_raw[, 1]
  tpm_mat <- as.matrix(tpm_raw[, -1])
  rownames(tpm_mat) <- tpm_ids
}

# Filter TPM matrix to retain the same filtered genes and matching sample order
tpm_filtered <- tpm_mat[rownames(dge$counts), metadata_scaled$sample]

# Save TPM table
tpm_df <- as_tibble(tpm_filtered, rownames = "gene_id")
tpm_df <- right_join(gene_annot, tpm_df, by = "gene_id")

# Write to Excel
exprfile <- file.path(resfolder, "TPM.xlsx")
write.xlsx(tpm_df, file = exprfile, colNames = TRUE, rowNames = FALSE)
```

Show the expression of these genes on a dot plot
```{r dotplot for selected genes}
hox_list <- list(
  "Cervical" = c("HOXA1", "HOXB1", "HOXA3", "HOXB3", "HOXD3"),
  "Truncal" = c("HOXA5", "HOXB5", "HOXA6", "HOXA7", "HOXB7", "HOXB8", "HOXC8", "HOXD8", "HOXB9"),
  "Lumbar-caudal" = c("HOXC9", "HOXD9", "HOXD11")
)
gene_order <- unlist(hox_list) 

# Convert your hox_list into a lookup table (dataframe)
hox_mapping <- stack(hox_list)
colnames(hox_mapping) <- c("gene_name", "gene_group")

# Ensure gene_group is a factor with the correct order for the facets
hox_mapping$gene_group <- factor(hox_mapping$gene_group, 
                                 levels = c("Cervical", "Truncal", "Lumbar-caudal"))

# Join this mapping back to your plotting data
exprmx_long_TPM <- tpm_df[, c("gene_name", metadata_scaled$sample)] %>%
  pivot_longer(cols = -gene_name, names_to = "sample", values_to = "gene_expr") %>%
  left_join(metadata_scaled, by = "sample")
exprmx_long_TPM <- exprmx_long_TPM %>%
  inner_join(hox_mapping, by = "gene_name")

# Crucial: Order the gene_name factor so they appear in sequence within the blocks
exprmx_long_TPM$gene_name <- factor(exprmx_long_TPM$gene_name, levels = gene_order)
# also order the samples
exprmx_long_TPM$sample <- factor(exprmx_long_TPM$sample, levels = unique(exprmx_long_TPM$sample))
exprmx_long_TPM$Target <- factor(exprmx_long_TPM$Target, levels = c("STOP_GFP", "STOP_MYCN"))

p_dot_samples <- ggplot(exprmx_long_TPM[exprmx_long_TPM$Target == "STOP_GFP", ], 
                        aes(x = gene_name, y = forcats::fct_rev(factor(sample)))) + 
  geom_point(aes(size = gene_expr, color = gene_expr)) +
  
  # Grouping into blocks
  facet_grid(. ~ gene_group, scales = "free_x", space = "free_x") +
  
  # Styling the color and size
  scale_color_gradient(low = "lightgrey", high = "red", name = "Average\nExpression") +
  scale_size_continuous(range = c(1, 8), name = "Average\nExpression") +
  guides(size = "none") +
  
  theme_bw() +
  labs(x = NULL, y = "Cell Type") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold.italic"),
    axis.text.y = element_text(size = 8),
    # Styling the facet labels (the "Blocks")
    strip.background = element_rect(fill = "gray90", color = "black"),
    strip.text = element_text(face = "bold", size = 10),
    panel.spacing = unit(0, "lines"), # Removes gaps between blocks for a seamless look
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.major = element_line(color = "gray95"),
    legend.position = "right"
  )

#print(p_dot_samples)
outfig <- file.path(figfolder, "selected_genes_tpm_dotplot_deltaonly.pdf")
ggsave(outfig, plot=p_dot_samples, width=10, height=10)

# all samples
p_dot_samples <- ggplot(exprmx_long_TPM, 
                        aes(x = gene_name, y = forcats::fct_rev(factor(sample)))) +
  # Size and Color both represent the expression level
  geom_point(aes(size = gene_expr, color = gene_expr)) +
  
  # Styling the color scale (Seurat-like red gradient)
  scale_color_gradient(low = "lightgrey", high = "red", name = "TPM") +
  
  # Adjust the dot size range
  scale_size_continuous(range = c(1, 8), name = "TPM") +
  guides(size = "none") +
  
  # Optional: Facet by condition to visually group your samples
  facet_grid(Target ~ ., scales = "free_y", space = "free_y") +
  
  theme_bw() +
  labs(x = "Genes", y = "Samples") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold.italic"),
    axis.text.y = element_text(size = 8),
    strip.text.y = element_text(angle = 0, face = "bold"), # Keep condition labels horizontal
    panel.grid.major = element_line(color = "gray95"),
    legend.position = "right"
  )
outfig <- file.path(figfolder, "selected_genes_tpm_dotplot.pdf")
ggsave(outfig, plot=p_dot_samples, width=10, height=10)


```

Add other samples from GEO
```{r read in other input data}
#knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

#gsenum <- "GSE172184"
vv_list <- list()
exprmx_list <- list()
samples_tmp_list <- list()
samples_orig_list <- list()
for (gsenum in c("GSE147635", "GSE94035")){
  infile <- here(paste("Data", gsenum, sep="_"),
                    "star_salmon/null.merged.gene.SummarizedExperiment.rds")
  gse_tmp <- readRDS(infile)
  colnames(gse_tmp) <- sub("_trimmed$", "", colnames(gse_tmp))
  
  annot_file_other <- here("Public_data", 
                               paste("SraRunTable_", gsenum, ".csv", sep=""))
  samples_annot_other <- read.csv(annot_file_other)
  table(samples_annot_other$tissue)
  #samples_tmp <- samples_annot[, c("Run", "tissue")]
  samples_tmp <- samples_annot_other
  #colnames(samples_tmp)[colnames(samples_tmp)=="Run"] <- "Sample"
  colnames(samples_tmp)[colnames(samples_tmp)=="tissue"] <- "Type"
  # remove missing data
  samples_tmp <- samples_tmp[samples_tmp$Run %in% colnames(gse_tmp),]
  # add all relevant information in the sample name
  if (gsenum == "GSE147635"){
    samples_tmp$Sample <- paste(samples_tmp$Sample.ID,
                              "DOD", samples_tmp$DOD,
                              "MYCN", samples_tmp$MYCN_amp, sep="_")
  }
  if (gsenum == "GSE94035"){
    samples_tmp$Sample <- paste(samples_tmp$patient,
                                samples_tmp$Tissue.extended, sep="_")
  }
  
  samples_tmp_list[[gsenum]] <- samples_tmp[, c("Sample", "Type")]
  samples_tmp_list[[gsenum]]$gse <- gsenum  
  samples_orig_list[[gsenum]] <- samples_tmp
  samples_orig_list[[gsenum]]$gse <- gsenum  

  # Option A: Additive Design Matrix
  design_gse <- model.matrix(~ Type, data = samples_tmp_list[[gsenum]])
  rownames(design_gse) <- samples_tmp_list[[gsenum]]$Sample
  
  # Extract the estimated counts and average effective lengths
  counts_tmp <- assay(gse_tmp, "salmon.merged.gene_counts")
  gene_lengths <- assay(gse_tmp, "salmon.merged.gene_lengths")
  
  # Convert estimated counts to a DGEList object (required by limma-voom)
  dge_tmp <- edgeR::DGEList(counts = counts_tmp)
  # change colnames to new names
  colnames(dge_tmp) <- samples_tmp$Sample[match(colnames(dge_tmp), samples_tmp$Run)]
  
  # Perform filtering to remove genes with low counts
  # A common threshold: Keep genes with CPM > 0.5 in at least two samples
  keep <- filterByExpr(dge_tmp, design_gse)
  dge_tmp <- dge_tmp[keep, , keep.lib.sizes=FALSE]
  
  # Apply the TMM normalization (a standard step in the limma-voom workflow)
  dge_tmp <- calcNormFactors(dge_tmp)
  
  # Apply voom normalization. This step calculates log2-CPM values, applies
  # the library size normalization (via the 'lib.size' in dge), and estimates
  # a precision weight for each observation, modeling the mean-variance relationship.
  vv_list[[gsenum]] <- voom(dge_tmp, design_gse, plot = F,
            # The 'lib.size' in DGEList is sufficient for Salmon counts,
            # but for the most precise limma-voom with Salmon data, we explicitly
            # pass the average transcript length matrix as an offset:
            # offset = log(gene_lengths[keep, ]) # (Uncomment this line for the most rigorous approach)
  )
  
  # save & plot intermediate results
  exprmx_list[[gsenum]] <- as_tibble(vv_list[[gsenum]]$E, rownames = "gene_id")
  print(dim(exprmx_list[[gsenum]]))
  # merge with gene annotation
  exprmx_list[[gsenum]] <- right_join(gene_annot, exprmx_list[[gsenum]])
  print(dim(exprmx_list[[gsenum]]))
}

# create a metadata with all information
metadata_all <- metadata_scaled[, c("sample", "Target")]
colnames(metadata_all)[which(colnames(metadata_all)=="sample")] <- "Sample"
colnames(metadata_all)[colnames(metadata_all)=="Target"] <- "Type"
metadata_all$gse <- "organoid"
samples_tmp <- Reduce(rbind, samples_tmp_list)
metadata_all <- rbind(metadata_all, samples_tmp)
# add malignancy to metadata
metadata_all$malignancy <- "malignant"
metadata_all$malignancy[metadata_all$Type=="delta_MYCN"] <- "benign"
metadata_all$malignancy[metadata_all$Type=="Ganglioneuroma"] <- "benign"
metadata_all$malignancy[metadata_all$Type=="Non-tumor mononuclear cells (MNCs)"] <- "benign"
metadata_all$Sample_lname <- paste(metadata_all$gse, metadata_all$Sample, sep=".")
# # save ?
# exprfile <- file.path(resfolder, "logCPM_gse.xlsx")
# write.xlsx(exprmx, file = exprfile, colNames=T, rowNames=F)
```

Read in GDC data
```{r read in data downloaded from GDC portal}
# first,  read in GDC datasheet
datasheet_file <- here("Data_GDC/gdc_sample_sheet.2025-11-13.tsv")
datasheet <- read_tsv(datasheet_file)
colnames(datasheet) <- make.names(colnames(datasheet))
# now add path to the actual files
datasheet <- datasheet %>%
  mutate(
    # Find the full path for each file name, searching recursively
    full_path = map_chr(File.Name, ~ {
      list.files(
        path = here("Data_GDC"), 
        pattern = paste0("^", .x, "$"), # Use regex to match exactly
        full.names = TRUE, 
        recursive = TRUE
      )
    })
  )
# for each sample, read in the counts data
counts_list <- list()
for (ID in datasheet$File.ID){
  ri <- which(datasheet$File.ID==ID)
  header <- read_tsv(datasheet$full_path[ri], skip=1, n_max=1)
   tmp  <- read_tsv(datasheet$full_path[ri], skip=6, col_names=F)
  colnames(tmp) <- colnames(header)
  counts_list[[ID]] <- tmp[, c("gene_id", "gene_name", "tpm_unstranded")]
  colnames(counts_list[[ID]])[colnames(counts_list[[ID]])=="tpm_unstranded"] <- ID
}
```


Analyse all samples together using limma and obtain a combined expression matrix (not only the gse samples)
```{r reanalyse with other input data}
#knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

# nf-core format: gene_id, gene_name, then one column per sample
organoid_counts <- as.matrix(counts_df[, !(colnames(counts_df) %in% c("gene_id", "gene_name"))])
rownames(organoid_counts) <- counts_df$gene_id

# Keep the same sample order as in the organoid metadata
organoid_counts <- organoid_counts[, metadata_scaled$sample]

expr_list_counts <- list("organoid"=organoid_counts)
for (gsenum in c("GSE147635", "GSE94035")) {
  infile <- here(paste("Data", gsenum, sep = "_"),
                 "star_salmon/null.merged.gene.SummarizedExperiment.rds")
  gse_tmp <- readRDS(infile)
  # use previously defined sample names
  colnames(gse_tmp) <- samples_orig_list[[gsenum]]$Sample[
    match(colnames(gse_tmp), samples_orig_list[[gsenum]]$Run)]

  # extract the count matrix right away
  expr_list_counts[[gsenum]] <- assay(gse_tmp, "salmon.merged.gene_counts")
}

# Check that all have identical gene rownames (they should)
all(sapply(expr_list_counts, 
    function(x) identical(rownames(x), rownames(expr_list_counts[[1]]))))
# if FALSE, you can do an inner join on rownames later

# Combine them by columns (samples)
expr_all_counts <- do.call(cbind, expr_list_counts)
dim(expr_all_counts)

# now provide a design matrix
metadata_all$Type <- as.factor(metadata_all$Type)
metadata_all$gse  <- as.factor(metadata_all$gse)
metadata_all$malignancy  <- as.factor(metadata_all$malignancy)
design_all  <- model.matrix(~ 0 + malignancy, data = metadata_all) # use this type of model matrix before as well???
#colnames(design_all) <- levels(metadata_all$Type)
rownames(design_all) <- metadata_all$Sample

# now the diff analysis and corrected expression
# Convert estimated counts to a DGEList object (required by limma-voom)
dge_all <- edgeR::DGEList(counts = expr_all_counts)

# Perform filtering to remove genes with low counts
# A common threshold: Keep genes with CPM > 0.5 in at least two samples
keep <- filterByExpr(dge_all, design_all)
dge_all <- dge_all[keep, , keep.lib.sizes=FALSE]

# Apply the TMM normalization (a standard step in the limma-voom workflow)
dge_all <- calcNormFactors(dge_all)

# Apply voom normalization. This step calculates log2-CPM values, applies
# the library size normalization (via the 'lib.size' in dge_all), and estimates
# a precision weight for each observation, modeling the mean-variance relationship.
vv_all <- voom(dge_all, design_all, plot = F,
          # The 'lib.size' in DGEList is sufficient for Salmon counts,
          # but for the most precise limma-voom with Salmon data, we explicitly
          # pass the average transcript length matrix as an offset:
          # offset = log(gene_lengths[keep, ]) # (Uncomment this line for the most rigorous approach)
)
```


Combine GDC counts data with all of the previous data
```{r combine GDC counts data with previous data}
# combine expression data with GDC data
expr_GDC <- counts_list %>% purrr::reduce(full_join, by = c("gene_id", "gene_name"))
#anyNA(expr_GDC)
expr_GDC$gene_name <- NULL
expr_GDC[, 2:ncol(expr_GDC)] <- log2(expr_GDC[, 2:ncol(expr_GDC)]+1)
expr_all_tbl <- as_tibble(vv_all$E, rownames="gene_id")

# harmonize ensembl ID-s
expr_GDC$gene_id_clean <- sub("\\..*", "", expr_GDC$gene_id)

mapping <- AnnotationDbi::select(
  org.Hs.eg.db,
  keys = expr_GDC$gene_id_clean,
  keytype = "ENSEMBL",
  columns = c("ENSEMBL", "SYMBOL")
)

expr_GDC$gene_current <- mapping$SYMBOL[match(expr_GDC$gene_id_clean, mapping$ENSEMBL)]

# remove NA rows
expr_GDC <- expr_GDC %>% drop_na(gene_current)
# remove columns that we don't need any more
# expr_GDC$gene_id <- NULL
expr_GDC$gene_current <- NULL # if biomart annotation
expr_GDC$gene_id_clean <- NULL

# expr_combined_tbl <- inner_join(expr_all_tbl, expr_GDC, by=c("gene_id"="gene_current")) 
expr_combined_tbl <- inner_join(expr_all_tbl, expr_GDC, by=c("gene_id"="gene_id"))   # if biomart annotation
expr_combined <- as.matrix(as.data.frame(expr_combined_tbl[, 2:ncol(expr_combined_tbl)]))
rownames(expr_combined) <- expr_combined_tbl$gene_id
  
metadata_GDC <- datasheet[, c("File.ID", "Tumor.Descriptor")]
colnames(metadata_GDC) <- c("Sample", "Type")
metadata_GDC$gse <- "TARGET-NBL"
metadata_GDC$malignancy <- "malignant"
metadata_GDC$Sample_lname <- metadata_GDC$Sample
metadata_combined <- bind_rows(metadata_all, metadata_GDC)

```

Calculate the differential gene expression for organoid data only and save the results.
```{r calculate differential gene expression}
fit <- lmFit(vv$E[, match(rownames(design_additive),colnames(vv$E))], 
             design=design_additive)
# save result for further use
outfile <- file.path(resfolder, "limma_fit_expr.rds")
saveRDS(fit, file=outfile)

# # also write to file
# gene_expr_average <- fit$coefficients
# colnames(gene_expr_average) <- samples_annot_short[match(
#               colnames(gene_expr_average), samples_annot_short$groupID), "group", drop=T]
# gene_expr_average <- as_tibble(gene_expr_average, rownames="Gene_ID")
# # put into a list
# exprout <- list()
# # add gene & sample annotation
# exprout[["mean"]]  <- genes_annot[match(gene_expr_average$Gene_ID,
#                                         genes_annot$Gene_ID),1:3]
# # add average gene expression
# exprout[["mean"]]  <- left_join(exprout[["mean"]] ,  
#                                 gene_expr_average, by="Gene_ID")
# outfile <- file.path(resfolder, "logCPM_mean.xlsx")
# write.xlsx(exprout[["mean"]] , file=outfile)

# contrasts 
contr.matrix <- makeContrasts(
  kolf2_vs_EpiC    = Cell_LineH2,
  STOP_vs_delta    = -TargetSTOP_GFP,   # STOP_MYCN - delta_MYCN
  gRNA2_vs_gRNA1   = gRNAgRNA2,
  levels = design_additive
)

fitc <- contrasts.fit(fit, contr.matrix)
fite <- eBayes(fitc) ## gives moderated t-statistics
# save result for further use
outfile <- file.path(resfolder, "limma_fit_diffexpr.rds")
saveRDS(fite, file=outfile)

# also write out to human readable file
lthr <- 1; pthr <- 0.05
thrchar <- paste(sprintf("lfc%01d", lthr), 
                   sprintf("adjp%.2f", pthr), sep="_")
outdata <- list()
outdata_signif <- list() # only write out genes whose diff expr is signif
for (cf in colnames(fite$coefficients)){
  toptt <- limma::topTable(fite, coef=cf, sort.by = "logFC",
          number=nrow(fite$coefficients))
  toptt <- as_tibble(toptt, rownames = "Gene_ID")
  # add annotation
  toptt <- left_join(toptt, gene_annot, by=c("Gene_ID"="gene_id"))
  lname <- substr(cf,1, min(nchar(cf), 31)) # for each tab name, a maximum of 31 length is allowed
  outdata[[lname]] <- toptt

  # repeat with significant genes only
  toptt_thr <- limma::topTable(fite, coef=cf, sort.by = "logFC",
                p.value=pthr, lfc=lthr, number=nrow(fite$coefficients))
  toptt_thr <- as_tibble(toptt_thr, rownames = "Gene_ID")
  # add annotation
  toptt_thr <- left_join(toptt_thr, gene_annot, by=c("Gene_ID"="gene_id"))
  outdata_signif[[lname]] <- toptt_thr
}
# write into a compound excel file
outfile <- file.path(resfolder, "diffexpr_all.xlsx")
write.xlsx(outdata, file = outfile, colNames=T, rowNames=F)
outfile_signif <- outfile <- file.path(resfolder, 
                     paste("diffexpr_", thrchar, "_signif.xlsx", sep=""))
write.xlsx(outdata_signif, file = outfile_signif, colNames=T, rowNames=F)

# calculate number of diff expressed genes for "kolf2_vs_EpiC"
table(outdata_signif[["kolf2_vs_EpiC"]]$seqnames)
```


Calculate the differential gene expression based on the extended count matrix
```{r calculate differential gene expression between ganglioneuroma vs STOP}

design_all_types  <- model.matrix(~ 0 + Type, data = metadata_all)
colnames(design_all_types) <- make.names(colnames(design_all_types))
rownames(design_all_types) <- metadata_all$Sample_lname
fit_all <- lmFit(vv_all$E[, match(rownames(design_all_types),colnames(vv_all$E))], 
             design=design_all_types)
# # save result for further use
# outfile <- file.path(resfolder, "limma_fit_expr.rds")
# saveRDS(fit, file=outfile)

# contrasts 
contr.matrix <- makeContrasts(
  STOP_minus_ganglioneuroma    =  TypeSTOP_MYCN-TypeGanglioneuroma,   # STOP_MYCN - delta_MYCN
  tumor_minus_ganglioneuroma_both = (TypePrimary.neuroblastoma.cell.line+TypeTumor.cells.from.the.primary.tumor+
                                 TypeDisseminated.tumor.cells..DTCs.+TypeTumor)/4 - TypeGanglioneuroma,
  tumor_minus_ganglioneuroma_GSE147635 = (TypePrimary.neuroblastoma.cell.line+
                                 TypeTumor.cells.from.the.primary.tumor)/2 - TypeGanglioneuroma,
  levels = colnames(design_all_types)
)

fitc_all <- contrasts.fit(fit_all, contr.matrix)
fite_all <- eBayes(fitc_all) ## gives moderated t-statistics
# # save result for further use
# outfile <- file.path(resfolder, "limma_fit_diffexpr.rds")
# saveRDS(fite, file=outfile)

# also write out to human readable file
lthr <- 1; pthr <- 0.05
thrchar <- paste(sprintf("lfc%01d", lthr), 
                   sprintf("adjp%.2f", pthr), sep="_")
outdata_all <- list()
outdata_all_signif <- list() # only write out genes whose diff expr is signif
for (cf in colnames(fite_all$coefficients)){
  toptt <- limma::topTable(fite_all, coef=cf, sort.by = "logFC",
          number=nrow(fite_all$coefficients))
  toptt <- as_tibble(toptt, rownames = "Gene_ID")
  # add annotation
  toptt <- left_join(toptt, gene_annot, by=c("Gene_ID"="gene_id"))
  lname <- substr(cf,1, min(nchar(cf), 31)) # for each tab name, a maximum of 31 length is allowed
  outdata_all[[lname]] <- toptt

  # repeat with significant genes only
  toptt_thr <- limma::topTable(fite_all, coef=cf, sort.by = "logFC",
                p.value=pthr, lfc=lthr, number=nrow(fite_all$coefficients))
  toptt_thr <- as_tibble(toptt_thr, rownames = "Gene_ID")
  # add annotation
  toptt_thr <- left_join(toptt_thr, gene_annot, by=c("Gene_ID"="gene_id"))
  outdata_all_signif[[lname]] <- toptt_thr
}
# write into a compound excel file
outfile <- file.path(resfolder, "diffexpr_ganglioneuroma_vs_STOP_MYCN.xlsx")
write.xlsx(outdata_all, file = outfile, colNames=T, rowNames=F)
outfile_signif <- outfile <- file.path(resfolder, 
                     paste("diffexpr_ganglioneuroma_vs_STOP_MYCN_", thrchar, "_signif.xlsx", sep=""))
write.xlsx(outdata_all_signif, file = outfile_signif, colNames=T, rowNames=F)

```




Heatmap using only STOP_MYCN, ganglioneuroma, neuroblastoma, neuroblastoma primary cell line and DTC samples
```{r plot heatmap with manually selected samples from GSE94035 and GSE147635}
  metadata_tmp1 <- metadata_combined[metadata_combined$gse %in% c("GSE147635", "organoid"),]
# additionally, from "GSE147635" keep all samples, from "GSE94035" keep non-relapsed MNC and DTC samples as well
  keep_GSE94035 <- c(grep("^GSE94035.*DTCs at diagnosis$", metadata_combined$Sample_lname),
                      grep("^GSE94035.*MNCs at diagnosis$", metadata_combined$Sample_lname))
  metadata_tmp2 <- metadata_combined[keep_GSE94035,]
  metadata_tmp <- bind_rows(metadata_tmp1, metadata_tmp2)
  hwidth=13; hheight=26
  cf_cur <- "tumor_minus_ganglioneuroma_GSE147635"
  
  # repeat with significant genes only
 lthr <- 5; pthr <- 0.001
 # use higher lthr (too many genes to plot)
 # lthr <- 5.5; pthr <- 0.001
 
 #cf = "STOP_minus_ganglioneuroma"
 #cf <- "tumor_minus_ganglioneuroma_both"
 ct <- "tumor_minus_ganglioneuroma_GSE147635"
  toptt_thr <- limma::topTable(fite_all, coef=cf, sort.by = "logFC",
                p.value=pthr, lfc=lthr, number=nrow(fite$coefficients))
  toptt_thr <- as_tibble(toptt_thr, rownames = "Gene_ID")
  # add annotation
  toptt_thr <- left_join(toptt_thr, gene_annot, by=c("Gene_ID"="gene_id"))
  # add MYCN
  MYCN_ID <- gene_annot$gene_id[gene_annot$gene_name=="MYCN"]
  
  # remove lncRNA-s
  toptt_thr <- toptt_thr %>% 
    filter(grepl("protein_coding", gene_type, ignore.case = TRUE))
  
  print("toptt dim:")
  print(dim(toptt_thr))

# repeat once more with more genes
# remove delta_MCYN samples
delta_MYCN_samples <- grep("STOP_GFP", metadata_tmp$Sample)
keep_samples <- setdiff(c(1:nrow(metadata_tmp)), delta_MYCN_samples)
#expr_heat_mx <- expr_combined[, colnames(expr_combined) %in% metadata_tmp$Sample_lname[keep_samples]]

expr_all <- as.matrix(as.data.frame(expr_all_tbl[, 2:ncol(expr_all_tbl)]))
rownames(expr_all) <- expr_all_tbl$gene_id
expr_heat_mx <- as.matrix(expr_all[, colnames(expr_all) %in% metadata_tmp$Sample_lname[keep_samples]])
dim(expr_heat_mx)
expr_heat_mx <- expr_heat_mx[rownames(expr_heat_mx) %in% c(toptt_thr$Gene_ID,  MYCN_ID),]
dim(expr_heat_mx)
rownames(expr_heat_mx) <- gene_annot$gene_name[match(rownames(expr_heat_mx), gene_annot$gene_id)]

# to add bottom bar:
annot_data <- metadata_tmp[match(colnames(expr_heat_mx), metadata_tmp$Sample_lname), ]
bottom_bar <- ComplexHeatmap::HeatmapAnnotation(
  Group = annot_data$Type, 
  col = list(Group = c(
    "Ganglioneuroma" = "black",
    "Non-tumor mononuclear cells (MNCs)" = "orange",
    "Disseminated tumor cells (DTCs)" = "yellow",
    "Tumor cells from the primary tumor" = "red",
    "STOP_MYCN" = "darkgreen",
    "Primary neuroblastoma cell line" = "blue"
  )),
  show_annotation_name = FALSE,
  show_legend = FALSE
)

# there would be too many genes to show the names - only show highlighted ones
key_genes <- c("MYCN", "AXL", "ROBO2", "HLA-DPB1") 
gene_indices <- match(key_genes, rownames(expr_heat_mx))
gene_indices <- na.omit(gene_indices)

# Create a right annotation with connector lines
gene_labels <- rowAnnotation(
  foo = anno_mark(
    at = gene_indices,
    labels = rownames(expr_heat_mx)[gene_indices],
    labels_gp = gpar(fontsize = 7, fontface = "bold.italic"),
    link_width = unit(5, "mm"),
    padding = unit(1, "mm")
  )
)

# select top differentially expressed genes
ht <- ComplexHeatmap::Heatmap(expr_heat_mx,
                        name = "logCPM\n",
                        clustering_method_rows = "complete",
                        row_names_gp = gpar(fontsize = 3),
                        #show_row_names = FALSE,
                        #right_annotation = gene_labels, # use these 2 rows if selected gene names are needed
                        column_names_gp = gpar(fontsize = 12),
                        column_names_max_height = unit(10, "cm"),
                        bottom_annotation = bottom_bar,
                        show_row_dend = FALSE,               #hide the left dendrogram
                        column_dend_height = unit(3, "cm"),  # enlarge the top dendrogram
                        # Pass a list of parameters to control the heatmap's main legend
    heatmap_legend_param = list(
        title_gp = gpar(
            fontsize = 26,     # Font size of the legend title
            fontface = "plain" # Font face of the legend title
        ),
        # Change the Legend Labels (Text):
        labels_gp = gpar(
            fontsize = 24       # Font size of the legend labels
        ),
        # Increase the size of the legend bar (for continuous legends)
        legend_height = unit(10, "cm")
    ))
#plot(ht)
tiff(file = file.path(figfolder, paste("heatmap_neuroblastoma_", "selected_samples", ".tiff", sep="")),
    width = hwidth, height = hheight, units = "in", res=300)
# pdf(file = file.path(figfolder, paste("heatmap_neuroblastoma_", "selected_samples", ".pdf", sep="")), 
#     width = hwidth, height = hheight)
draw(ht)
dev.off()

```




Volcano plot for STOP vs delta
```{r MYCN volcano plot}
# needs to be added later: highlight some genes
pthr <- 0.05
lthr <- 1
toptt_MYCN <- limma::topTable(fite, coef="STOP_vs_delta", sort.by = "none",
        number=nrow(fite$coefficients))
toptt_MYCN <- as_tibble(toptt_MYCN, rownames = "Gene_ID")
toptt_MYCN <- left_join(toptt_MYCN, gene_annot, by=c("Gene_ID"="gene_id"))


    # add whether it was differentially expressed
toptt_MYCN$diffexpressed <- "NO"
toptt_MYCN$diffexpressed[toptt_MYCN$logFC > lthr  & toptt_MYCN$adj.P.Val < pthr ] <- "UP"
toptt_MYCN$diffexpressed[toptt_MYCN$logFC < -lthr  & toptt_MYCN$adj.P.Val < pthr ] <- "DOWN"
toptt_MYCN$diffexpressed <- factor(toptt_MYCN$diffexpressed, levels=c("NO", "UP", "DOWN"))

# # add text for selected & top genes
# selected_genes <- c("ABCC1", "ABCC3", "ABCG2", "CDK1", "CDK2", "CDK4", "CDK5", 
#     "CDK7", "CDK9", "DNMT1", "DNMT3A", "DNMT3B", "DCK", "FGFR1", "KIT", "PDGFRA", 
#     "RRM1", "RRM2", "RRM2B", "SAMHD1","FLT1", "FDPS", "XPO1", "IDH1", "IDH2", "BCL2", "PLK1",
#     "TOP2A", "NEDD8", "MDM2", "ABCB1", "BTK", "EGFR", "FLT3", "FTO", "PSMB5")
# keep_genes <- genes_annot$Gene_ID[match(selected_genes, 
#                                  genes_annot$name)]
# keep_ind <- match(keep_genes, rownames(toptt_HFF))
keep_ind <- c()
# also the "large" differential expression ones
# lfc_spec <- 7.5
# adjp_spec <- 2
lfc_spec <- 15
adjp_spec <- 10
large_genes_ind <- which(abs(toptt_MYCN$logFC)>=lfc_spec |
                       -log10(toptt_MYCN$adj.P.Val)>=adjp_spec)
large_genes_names <- toptt_MYCN$gene_name[large_genes_ind]
# create a small dataframe with the labels
text_df <- toptt_MYCN[c(keep_ind, large_genes_ind), ]


# p <- ggplot(data=toptt_HFF, aes(x=logFC, y=-log10(adj.P.Val), 
#                                  col=diffexpressed)) +
p <- ggplot(data=toptt_MYCN, aes(x=logFC, y=-log10(adj.P.Val), 
                                 col=diffexpressed)) +
              geom_point() + 
              geom_point(data = subset(toptt_MYCN, gene_name == "MYCN"),    
                         shape = 21,          # filled circle with outline
                        fill = "red",      # red center (or "black" for filled)
                        color = "black",     # black border
                        stroke = 1.2         # line thickness of the border
                      ) +
              theme_minimal() +
              scale_color_manual(name="differentially\n expressed",
                values=c("DOWN"="blue", "NO"="gray", "UP"="red")) +
              geom_vline(xintercept=c(-lthr, lthr), col="black", linetype="dashed") +
              geom_hline(yintercept=-log10(pthr), col="black", linetype="dashed")+
              # geom_text(data=text_df, aes(x=logFC, y=-log10(adj.P.Val),
              #                             label=gene_name), color="black",
              #            size=2, nudge_y = 0.05)+
                geom_text_repel(
                  data = text_df,
                  aes(label = gene_name),
                  size = 2,
                  box.padding = 0.4,     # distance between text and points
                  point.padding = 0.2,   # distance between text and its point
                  segment.color = "gray50",  # connecting line color
                  max.overlaps = Inf     # don’t drop labels
                ) +
                geom_text_repel(
                  data = subset(toptt_MYCN, gene_name == "MYCN"),
                  aes(label = gene_name),
                  size = 4,
                  color = "black",
                  fontface = "bold",
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = "black"
                ) +
              theme_gray(base_size = 20)+
              #xlab(bquote(Log[2] * bgroup("(", "Fold Change", ")" )))+
              #ylab(bquote(-Log[10] * bgroup("(", "adjusted P value", ")" )))
              xlab(bquote(log[2] ~ (fold ~ change)))+
              ylab(bquote(-log[10] ~ (adjusted ~ p ~ value)))
outfig <- file.path(figfolder, "volcano_geneset_MYCN.pdf")
ggsave(p, file=outfig, width=10, height=7.5)
outfig <- file.path(figfolder, "volcano_geneset_MYCN.tiff")
# ggsave(p, file=outfig, width=10, height=7.5)

```

Create a list of significantly up- and downregulated genes.
```{r create a list of siginficantly up- and downregulated genes}
lthr <- 1; pthr <- 0.05
signif_diff <- list()
for (cf in colnames(fite$contrasts)){
  toptt <- topTable(fite, coef=cf,  
                    number=nrow(fite$coefficients), sort.by = "none",
                    p.value=pthr, lfc = lthr)
  name_up <- paste(cf,"up", sep="_")
  signif_diff[[name_up]] <- rownames(toptt)[toptt$logFC>0]
  name_down <- paste(cf, "down", sep="_")
  signif_diff[[name_down]] <- rownames(toptt)[toptt$logFC<0]
}
# number of genes
setsizes <- unlist(lapply(signif_diff, length))
```



## Functional Enrichment Analysis

To interpret the biological significance of the identified differentially expressed (DGE) and equivalent gene sets, we performed functional enrichment analysis using curated gene set collections from the Molecular Signatures Database (MSigDB).

### Gene Set Collections
We utilized three distinct categories of gene sets to capture a broad range of biological insights:

* **Hallmark Gene Sets:** Represent specific, well-defined biological states or processes (e.g., hypoxia, inflammatory response) and provide a concise summary of the primary pathways involved.
* **KEGG Medicus:** A subset of the Kyoto Encyclopedia of Genes and Genomes (KEGG) that focuses on the molecular networks of diseases and drug mechanisms, offering a more clinical and mechanistic perspective.
* **GO Biological Process (GO:BP):** Comprehensive ontologies that describe the broader biological objectives (such as cell cycle or metabolic pathways) to which a gene contributes.

This multi-faceted approach allows us to determine if the genes identified as "changing" or "stable" are non-randomly associated with specific biological functions or signaling pathways.

```{r functional analysis, message=FALSE, warning=FALSE}
geneset_collections <- list(
  Hallmark = msigdbr(species = "Homo sapiens", collection = "H") %>%
             dplyr::select(gs_name, gene_symbol, ensembl_gene),

  KEGG_Medicus = msigdbr(species = "Homo sapiens", collection = "C2") %>%
                 filter(gs_subcollection == "CP:KEGG_MEDICUS") %>%
                 dplyr::select(gs_name, gene_symbol, ensembl_gene),

  GO_Biological_Process = msigdbr(species = "Homo sapiens", collection = "C5") %>%
                          filter(gs_subcollection == "GO:BP") %>%
                          dplyr::select(gs_name, gene_symbol, ensembl_gene)
)
```

# Create gene annotation
```{r}
# 1. Get your list of gene symbols from the object
current_symbols <- rownames(seu.integrated)

# 2. Query the database for IDs/Annotations
gene_info <- select(org.Hs.eg.db,
                    keys = current_symbols,
                    columns = c("ENSEMBL", "GENENAME"),
                    keytype = "SYMBOL")

# 3. Clean up the query (handle potential duplicates/NAs)
# (Optional: remove duplicates if a symbol maps to multiple Ensembl IDs)
gene_info <- gene_info[!duplicated(gene_info$SYMBOL), ]
# keep non-NA only
gene_info <- gene_info[!is.na(gene_info$ENSEMBL),]
```

### Gene Set Enrichment Analysis (GSEA)

While Overrepresentation Analysis (ORA) treats gene lists as unranked sets, **Gene Set Enrichment Analysis (GSEA)** utilizes the full spectrum of gene expression changes. This allows for the detection of subtle but coordinated shifts in entire biological pathways that might be missed by simple threshold-based filtering.

#### 1. Analysis Workflow: Ranking and Running
For each experimental contrast, we performed the following steps:
* **Gene Ranking:** All genes were ranked by their $log_2(\text{Fold Change})$ in descending order. This ensures that the most upregulated genes are at the "top" of the list and the most downregulated genes are at the "bottom."
* **Enrichment Calculation:** Using the `clusterProfiler::GSEA` function, we calculated a running sum statistic to identify pathways where genes are non-randomly distributed at either end of the ranked list.

#### 2. Key Metrics and "Core Enrichment"
The results are summarized using two primary metrics:
* **Normalized Enrichment Score (NES):** A positive NES indicates the pathway is generally **activated** (enriched in upregulated genes), while a negative NES indicates the pathway is **suppressed** (enriched in downregulated genes).
* **Core Enrichment:** We extracted the specific subset of "leading-edge" genes—those that contribute most significantly to the enrichment score—and reported them along with their individual $log_2(\text{Fold Change})$ values.

#### 3. Visualization and Export
* **Summary Dotplots:** For each comparison, we visualized the top 10 activated and top 10 suppressed pathways. The size of the points represents the significance ($-\log_{10}(\text{Adjusted P-Value})$), and the x-axis indicates the NES.

#### Output file
Results are saved to GSEA_differential_expression_[geneset_name]_results.xlsx file, where each tab contains the results of one of the contrasts and only those genesets are listed for each contrast where the adjusted p value <= 0.05. If the enrichment process resulted no genesets, then the contrast is not shown in the file. If none of the genesets were significant, then the corresponding tab is empty.

After Calculate the differential gene expression for organoid data only and save the results.
```{r GSEA for differentially expressed genes, message=FALSE, warning=FALSE, fig.width=12, fig.height=12}
gsea_outer_list <- list()
p_multi_list <- list()
 for (gs_name in names(geneset_collections)){
   current_t2g <- geneset_collections[[gs_name]]
   
    # add gene symbols
    id_map <- current_t2g %>%
    dplyr::select(ensembl_gene, gene_symbol) %>%
    distinct() %>%
      deframe() # Converts to a named vector
   
   # print("########################################")
   # print(paste("#   Current gene set:", gs_name, "    #"))
   # print("########################################")
   gsea_res_list <- list()
   # Iterate through your contrasts
   # for (contr in colnames(fite$coefficients)) {
   for (contr in c("STOP_vs_delta")) {
     # print(paste("Current contrast:", contr))

          # 1. Prepare the ranked list for GSEA
     full_stats <- topTable(fite, coef = contr, number = Inf)

     # Create a named vector for GSEA ranking
     gene_list <- full_stats$logFC
     names(gene_list) <- sub("\\..*$", "", rownames(full_stats))

     # Create a lookup for the table mapping (clean Ensembl IDs)
     lfc_lookup <- full_stats$logFC
     names(lfc_lookup) <- sub("\\..*$", "", rownames(full_stats))

     # Sort for GSEA
     gene_list <- sort(gene_list, decreasing = TRUE)

     # 2. Run GSEA
     gsea_out <- clusterProfiler::GSEA(
       geneList = gene_list,
       TERM2GENE = current_t2g[, c("gs_name", "ensembl_gene")],
       pvalueCutoff = 0.05,
       verbose = FALSE
     )

     if (is.null(gsea_out) || nrow(as.data.frame(gsea_out)) == 0) {
       next
     }

     # 3. Process results and map Symbols + LogFC
     result_df <- as.data.frame(gsea_out)

     # 4. Update the mapping logic to include LogFC for Core Enrichment
     result_df_final <- result_df %>%
       mutate(MSigDB_Link = paste0("<a href='https://www.gsea-msigdb.org/gsea/msigdb/cards/",
                                   Description, "' target='_blank'>Link</a>")) %>%
       mutate(core_enrichment_stats = map_chr(core_enrichment, function(id_string) {
         # Split IDs (GSEA uses "/" as delimiter)
         ids <- unlist(strsplit(id_string, "/"))

         # Get Symbols from your id_map
         symbols <- id_map[ids]
         symbols[is.na(symbols)] <- ids[is.na(symbols)] # fallback to ID if symbol missing

         # Get LogFC values from our lookup
         logfcs <- lfc_lookup[ids]

         # Format as "Symbol (LogFC)"
         formatted <- paste0(symbols, " (", round(logfcs, 2), ")")

         # Combine into a single string
         paste(formatted, collapse = ", ")
       })) %>%
       # Select columns for the final table
       dplyr::select(Description, NES, p.adjust, core_enrichment_stats, MSigDB_Link)

     # Format p-value to scientific
     result_df_final$p.adjust <- formatC(as.numeric(result_df_final$p.adjust), format = "e", digits = 2)
     result_df_final$NES <- formatC(as.numeric(result_df_final$NES), format = "e", digits = 2)

     # Add to list for Excel export
     gsea_res_list[[contr]] <- result_df_final

     # also add a figure
     if (!is.null(gsea_out) && nrow(as.data.frame(gsea_out)) > 0) {

     # 1. Identify the top 3 pathways by absolute NES
     # (You can change this to show the most significant or specific pathways of interest)
     top_pathways <- as.data.frame(gsea_out) %>%
       arrange(desc(abs(NES))) %>%
       slice_head(n = 3) %>%
       pull(ID)

     # # 2. Generate the plot for the top pathway
     # # p1: The Enrichment Score curve
     # # p2: The gene 'ticks' (hits)
     # # p3: The rank-ordered gene list gradient
     # p_gsea <- gseaplot2(gsea_out,
     #                     geneSetID = top_pathways[1],
     #                     title = paste("Top Pathway:", top_pathways[1]),
     #                     subplots = 1:3,
     #                     pvalue_table = TRUE)
     # 
     # # 3. Save the plot as a high-resolution PDF or PNG
     # # Since gseaplot2 returns a ggplot-compatible object, we use ggsave
     # plot_filename <- file.path(figfolder, paste0("GSEA_Plot_", contr, ".pdf"))
     # 
     # ggsave(plot_filename,
     #        plot = p_gsea,
     #        width = 10,
     #        height = 8,
     #        dpi = 300,
     #        bg = "white")

     # Optional: Plot multiple top pathways together
     p_multi <- enrichplot::gseaplot2(gsea_out,
                          geneSetID = top_pathways,
                          title = paste("Top 3 Pathways in", contr))
     p_multi_wrapped <- patchwork::wrap_plots(p_multi, ncol = 1, heights = c(2, 0.5, 0.5))
     # p_multi_list[[contr]] <- p_multi_wrapped
     subfolder <- file.path(figfolder, gs_name)
     if (!file.exists(subfolder)) dir.create(subfolder)
     ggsave(file.path(subfolder, paste0("GSEA_MultiPlot_", contr, ".pdf")),
            plot = p_multi, width = 12, height = 9)
     }
   }

   # Write out to Excel
   outfile_gsea <- file.path(resfolder, paste("GSEA_differential_expression_", gs_name, "_results.xlsx", sep=""))
   write.xlsx(gsea_res_list, file = outfile_gsea, colNames = TRUE)
   gsea_outer_list[[gs_name]] <- gsea_res_list
 }
```


```{r dotplot of GSEA results, fig.width=20, fig.height=20}
for (gs_name in names(gsea_outer_list)){
  gsea_res_list <- gsea_outer_list[[gs_name]]
  plist <- list()
  for (gsi in names(gsea_res_list)){
    # create nice title:
    gsi_title <- switch(gsi,
      "TumorVsNormal"="Tumor vs Normal",
      "STOP_vs_delta" = "STOP vs delta",
      gsi
    )
    clean_title <- paste(gs_name, gsi_title, sep = " - ")
    clean_title <- gsub("_", " ", clean_title)
    
    y <- gsea_res_list[[gsi]]
    
    # select activated and supressed genesets
    y$NES <- as.numeric(y$NES)
    y$p.adjust <- as.numeric(y$p.adjust)
    y$direction <- "activated"
    y$direction[y$NES<=0] <- "suppressed"
    numact <- length(which(y$direction=="activated"))
    numsupp <- length(which(y$direction=="suppressed"))

    # pick top and bottom genesets
    act_Nvals <- y$NES[y$direction=="activated"]
    act_thr <- sort(act_Nvals, decreasing = T)[min(15, numact)]
    act_keep <- which(y$direction=="activated" & y$NES >= act_thr)
    sup_Nvals <- y$NES[y$direction=="suppressed"]
    sup_thr <- sort(sup_Nvals, decreasing = F)[min(20, numsupp)]
    sup_keep <- which(y$direction=="suppressed" & y$NES <= sup_thr)
    y <- y[c(act_keep, sup_keep),]
    
    # sort by normalized enrichment
    y <- y[order(y$NES, decreasing = F),]
    # change the names of the gene sets
    #y$Description <- tolower(y$Description)
    #y$Description <- gsub(prefix[gsi], "", y$Description) # later if needed prefix <- c("gobp_", "kegg_", "hallmark_")
    y$Description <- gsub("_", " ", y$Description)
    y$Description <- str_to_sentence(y$Description)
    y$Description <- gsub("Gobp", "GO:BP", y$Description)
    y$Description <- gsub("dna", "DNA", y$Description)
    y$Description <- gsub("Dna", "DNA", y$Description)
    y$Description <- str_wrap(y$Description, width = 50) # width=30 is a good starting point for a 4-column layout
    #y$Description <- nice_names$name[match(y$Description, nice_names$compare)]
    y$size <- -log10(as.numeric(y$p.adjust))
    
    num_suppressed <- sum(y$direction == "suppressed")
    sep_line_pos <- num_suppressed + 0.5
    y$Description <- factor(y$Description, levels=y$Description)
      p <- ggplot(y, 
                    aes(x = NES, y = Description)) + 
               geom_point(aes(size = size, color = direction))+
               scale_color_manual(values = c("activated" = "firebrick", "suppressed" = "dodgerblue4")) +
               
              scale_size_continuous(
                 name = "-log10(p.adj)", 
                 limits = c(0, 20),      # <--- This is the fixed range
                 breaks = c(2, 4, 6, 8) # <--- Fixed circle sizes in legend
               ) +  
              
              # Only add it if both directions exist to avoid errors
             {if(num_suppressed > 0 && num_suppressed < nrow(y)) 
               geom_hline(yintercept = sep_line_pos, linetype = "solid", color = "gray80", size = 0.5)} +
              
               guides(size=guide_legend(title="-log(p.adjust)", ncol=1,byrow=TRUE,
                                title.vjust=0.9, order=1))+
               theme_bw(base_size = 16) +

               theme(
                 axis.text.y = element_text(size = 16, lineheight = 0.8),
                 panel.grid.minor = element_blank(),
                 legend.position="right",
                      legend.box = "vertical",
                     #axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                     #strip.background =element_rect(fill=cfill),
                     strip.text = element_text(colour = 'black'),
                 legend.title = element_text(size = 16), # Legend title size
                  legend.text = element_text(size = 14)                 # Legend item text size)
               )+
               ylab(NULL) + xlab("NES")+
               #scale_size(range=c(3, 8))+
               #scale_y_discrete(labels = function(x) str_wrap(x, width = 24))+
               geom_vline(xintercept = 0, linetype="dashed")+
               labs(title=clean_title)
      #plot(p)
      plist[[gsi]] <- p
    # figfile <- file.path(figfolder, paste("GSEA_", gs_name, "_", gsi, "_enrich_dotplot.pdf", sep=""))
    # ggsave(figfile, plot=p, width=14, height=6)
  }
  # plot together for the three genesets
  pp_combined <- patchwork::wrap_plots(plist, ncol = 1) + 
  patchwork::plot_layout(guides = "collect") & # Collects legends into one place
  theme(legend.position = "right")
  # plot(pp_combined)
  ggsave(file.path(figfolder, paste("GSEA_", gs_name, "_enrich_dotplot_combined.pdf", sep="")), 
         plot = pp_combined, width = 10, height = 10)
}
```
